---
title: "Utah Prison Population Projections"
subtitle: "Historical Trends and Future Projections"
format: 
  dashboard:
    orientation: rows  
    logo: images/CCJJ_Logo_1.png
    theme: 
      - sandstone
      - custom.scss
    nav-buttons:
      - linkedin
      - twitter
      - icon: gitlab
        href: https://gitlab.com/
    scrolling: false
---

```{r}
#| label: setup
#| message: false
#| include: false

# Load required packages - streamlined to only what's needed for visualization
library(tidyverse)
library(plotly)
library(DT)
library(scales)
library(bslib)
library(bsicons)
library(lubridate)
library(kableExtra)
library(knitr)
library(readxl)
library(dplyr)
library(lubridate)
library(tsibble)
library(ggplot2)
library(plotly)
library(scales)
library(DT)
library(forecast)
library(htmltools)  # For tags$p function
library(fable)

# Set global options
options(scipen = 999)
theme_set(theme_minimal())

# Define parameters that would have been selected by the user
population_type <- "total"  # Options: "total", "male", "female"
projection_horizon <- 20    # Options: 8, 12, 20
scenario <- "baseline"      # Options: "baseline", "reform", "covid"
```

```{r}
#| label: load-preprocessed-data
#| message: false

# Load pre-processed data files
prison_data_clean <- read.csv("data/prison_data_clean.csv") %>%
  mutate(Quarter_Year = yearquarter(Quarter_Year))

# Load ADJUSTED forecast dataframes
forecast_df <- read.csv("data/forecast_total_adjusted.csv") %>%
  mutate(Quarter_Year = yearquarter(Quarter_Year))

forecast_df_m <- read.csv("data/forecast_male_adjusted.csv") %>%
  mutate(Quarter_Year = yearquarter(Quarter_Year)) %>%
  # Use the adjusted forecast column as the main forecast column for consistency
  mutate(Forecasted = Forecasted_male_adj,
         Lower_CI = Lower_CI_adj,
         Upper_CI = Upper_CI_adj)

# Female forecast remains unchanged (no adjustment was applied)
forecast_df_f <- read.csv("data/forecast_female.csv") %>%
  mutate(Quarter_Year = yearquarter(Quarter_Year))

# Load adjusted forecast tables
forecast_table_hybrid <- read.csv("data/forecast_table_total_adjusted.csv") %>%
  mutate(Quarter_Year = yearquarter(Quarter_Year))

# Load combined forecast data for male/female tables
forecast_combined <- read.csv("data/forecast_combined_adjusted.csv") %>%
  mutate(Quarter_Year = yearquarter(Quarter_Year))

# Extract forecast table for males from the combined data
forecast_table_hybrid_m <- forecast_combined %>%
  filter(is.na(Actual_male)) %>%
  select(Quarter_Year, Forecasted_male_adj) %>%
  rename(Forecasted_hybrid = Forecasted_male_adj) %>%
  mutate(
    Percent_Change = round((Forecasted_hybrid - lag(Forecasted_hybrid)) / lag(Forecasted_hybrid) * 100, 1)
  )

# Female forecast table remains unchanged
forecast_table_hybrid_f <- read.csv("data/forecast_table_female.csv") %>%
  mutate(Quarter_Year = yearquarter(Quarter_Year))

# Load summary metrics (these already contain the adjusted values)
summary_metrics <- read.csv("data/summary_metrics.csv")

# Extract current metrics from summary data
current_total <- summary_metrics %>% filter(population_type == "total") %>% pull(current_pop)
current_male <- summary_metrics %>% filter(population_type == "male") %>% pull(current_pop)
current_female <- summary_metrics %>% filter(population_type == "female") %>% pull(current_pop)

# Extract projected metrics (these are the adjusted values)
projected_total <- summary_metrics %>% filter(population_type == "total") %>% pull(projected_pop)
projected_male <- summary_metrics %>% filter(population_type == "male") %>% pull(projected_pop)
projected_female <- summary_metrics %>% filter(population_type == "female") %>% pull(projected_pop)

# Extract percent changes (these are the adjusted values)
pct_change_total <- summary_metrics %>% filter(population_type == "total") %>% pull(pct_change)
pct_change_male <- summary_metrics %>% filter(population_type == "male") %>% pull(pct_change)
pct_change_female <- summary_metrics %>% filter(population_type == "female") %>% pull(pct_change)

# Extract annual changes (these are the adjusted values)
annual_change_total <- summary_metrics %>% filter(population_type == "total") %>% pull(annual_change)
annual_change_male <- summary_metrics %>% filter(population_type == "male") %>% pull(annual_change)
annual_change_female <- summary_metrics %>% filter(population_type == "female") %>% pull(annual_change)



# Read in Prison Admission Data
admin_rel_merged_df <- read.csv("data/admin_rel_merged_df.csv")

# Load capacity data
capacity_data <- read.csv("data/capacity_data.csv")
capacity_limits <- read.csv("data/capacity_limits.csv")

# Load proportion data (this uses adjusted values)
proportion_data <- read.csv("data/proportion_data.csv") %>%
  mutate(Quarter_Year = yearquarter(Quarter_Year))


# Load gender & offense type data for faceted plots
df_driver_1 <- read.csv("data/df_driver_1.csv") %>%
  mutate(Quarter_Year = yearquarter(Quarter_Year))

df_driver_2 <- read.csv("data/df_driver_2.csv") %>%
  mutate(Quarter_Year = yearquarter(Quarter_Year))

df_driver_3 <- read.csv("data/df_driver_3.csv") %>%
  mutate(Quarter_Year = yearquarter(Quarter_Year))


# Helper function to select appropriate data based on population type
get_data_for_population <- function(pop_type = population_type) {
  # Get capacity data for the specific population type
  capacity_limit <- capacity_data %>% 
    filter(type == pop_type) %>%
    pull(capacity)
  
  # Get the detailed capacity limits if available
  capacity_limits_detail <- capacity_limits %>% 
    filter(type == pop_type)
  
  # Set default values in case the detailed capacity data isn't available
  absolute_max <- capacity_limit
  emergency <- capacity_limit * 0.98  # Default fallback if detailed data not found
  operational <- capacity_limit * 0.96  # Default fallback if detailed data not found
  
  # Use detailed capacity values if available
  if (nrow(capacity_limits_detail) > 0 && "capacity_type" %in% colnames(capacity_limits_detail)) {
    absolute_max <- capacity_limits_detail %>% 
      filter(capacity_type == "absolute_max") %>% 
      pull(capacity)
    
    emergency <- capacity_limits_detail %>% 
      filter(capacity_type == "emergency") %>% 
      pull(capacity)
    
    operational <- capacity_limits_detail %>% 
      filter(capacity_type == "operational") %>% 
      pull(capacity)
  }
  
  if (pop_type == "total") {
    list(
      forecast_data = forecast_df,
      table_data = forecast_table_hybrid,
      current_pop = current_total,
      projected_pop = projected_total,
      pct_change = pct_change_total,
      annual_change = annual_change_total,
      title = "Total Prison Population",
      color = "#00A7C4",
      # Return both individual capacity values and the full dataframe
      capacity = absolute_max,  # For backward compatibility with existing code
      capacity_absolute_max = absolute_max,
      capacity_emergency = emergency,
      capacity_operational = operational,
      capacity_data = capacity_limits_detail
    )
  } else if (pop_type == "male") {
    list(
      forecast_data = forecast_df_m,
      table_data = forecast_table_hybrid_m,
      current_pop = current_male,
      projected_pop = projected_male,
      pct_change = pct_change_male,
      annual_change = annual_change_male,
      title = "Male Prison Population",
      color = "#0a4c6a",
      # Return both individual capacity values and the full dataframe
      capacity = absolute_max,  # For backward compatibility with existing code
      capacity_absolute_max = absolute_max,
      capacity_emergency = emergency,
      capacity_operational = operational,
      capacity_data = capacity_limits_detail
    )
  } else {
    list(
      forecast_data = forecast_df_f,
      table_data = forecast_table_hybrid_f,
      current_pop = current_female,
      projected_pop = projected_female,
      pct_change = pct_change_female,
      annual_change = annual_change_female,
      title = "Female Prison Population",
      color = "#fc719e",
      # Return both individual capacity values and the full dataframe
      capacity = absolute_max,  # For backward compatibility with existing code
      capacity_absolute_max = absolute_max,
      capacity_emergency = emergency,
      capacity_operational = operational,
      capacity_data = capacity_limits_detail
    )
  }
}


# Get data for the currently selected population type
selected_data <- get_data_for_population(population_type)
```

#  {.sidebar}

### About
<span style="color: #1696d2">*The Utah Prison Projection Project is a collaborative effort between the Utah Department of Corrections and the Commission on Criminal and Juvenile Justice. This initiative aims to develop a Utah-specific prison population forecasting tool to support criminal justice resource planning and analyze the impact of new legislation and policy on prison populations.* </span>

<span style="color: #565656"> **This dashboard provides projections of Utah's prison population based on historical data.**</span>

<span style="color: #565656"> **Population Type:** Total, Male, Female </span>

<span style="color: #565656">**Projection Horizon:** 5-years </span>

<span style="color: #565656">**Policy Scenario:** Baseline </span>

<span style="color: #565656">**Capacity:** Absolute Maximum Capacity (100% threshold), Emergency Capacity (98% threshold), Operational Capacity (96.5% threshold) </span>



# 🏛️ Overview {orientation="rows"}

## Value Boxes {height="40%"}

```{r}
#| label: value-boxes

# Get data based on the selected population type
data <- get_data_for_population(population_type)

# Create layout with value boxes
layout_column_wrap(
  width = "250px",
  value_box(
    title = paste(str_to_title(gsub("Population", "", population_type)), "Current Population"),
    value = comma(round(data$current_pop)),
    showcase = bs_icon("people-fill"),
    theme = value_box_theme(bg = "#e6f2fd", fg = "#0B538E"),
    tags$p("2024 Q4")
  ),
  value_box(
    title = paste(str_to_title(gsub("Population", "", population_type)), "Projected Population"),
    value = comma(round(data$projected_pop)),
    showcase = bs_icon("graph-up"),
    theme = "teal",
    tags$p("5 years projection"),
    tags$p("2029 Q4")
  ),
  value_box(
    title = "Percent Change",
    value = paste0(data$pct_change, "%"),
    showcase = bs_icon("percent"),
    theme = "pink",
    tags$p("over 5 years")
  ),
  value_box(
    title = "Average Annual Change",
    value = paste0(data$annual_change, "%"),
    showcase = bs_icon("calendar-event"),
    theme = "warning",
    tags$p("annually")
  )
)
```

## Plot {height="60%"}

```{r}
#| label: admissions-releases-plot
#| title: Quarterly Net Difference Prison Admissions and Releases Over Time
#| height: 60%

# Plot difference
plot_ly(admin_rel_merged_df, x = ~as.character(Year_Quarter)) %>%
  add_lines(y = ~diff, name = "Net difference", line = list(color = "#00A7C4")) %>%
  layout(
    title = "",
    xaxis = list(title = "Year-Quarter",  
                 zerolinecolor = "#fdbf11",  # Yellow zero line
                 zerolinewidth = 2           # Make the zero line more visible
    ),
    yaxis = list(title = "Count", 
                 range = c(0, NA),
                 zerolinecolor = "#fdbf11",  # Yellow zero line for y-axis too
                 zerolinewidth = 2
                 ),  # Start y-axis at 0
     plot_bgcolor = "#f5f5f5",     # Light grey plot background
    paper_bgcolor = "#ffffff"     # White paper background
  )
```

# 📈 Projections {.tabset}

## Total Population

### Column {width=70%}

```{r}
#| label: total-projection-plot

# Get the appropriate forecast data for total population
total_data <- get_data_for_population("total")
forecast_data <- total_data$forecast_data
color_actual <- total_data$color
title <- paste(total_data$title, "Projection")

# Find the transition point between actual and forecast
actual_data <- forecast_data[!is.na(forecast_data$Actual),]
max_date <- max(actual_data$Quarter_Year)

# Get horizon dates for filtering
horizon_dates <- seq(max_date + 1, by = 1, length.out = projection_horizon)

# Create plot
p <- plot_ly(
  hoverlabel = list(namelength = -1)  # Don't truncate trace names in hover
)

# Add historical data
p <- p %>% add_trace(
  data = forecast_data[!is.na(forecast_data$Actual),],
  x = ~Quarter_Year,
  y = ~Actual,
  type = 'scatter',
  mode = 'lines',
  name = "Historical",
  line = list(color = color_actual),
  hoverinfo = "text",
  text = ~paste0(
    format(Quarter_Year, "%Y Q"), ceiling(as.numeric(format(Quarter_Year, "%m"))/3),
    "<br>Population: ", formatC(Actual, format="d", big.mark=",")
  )
)

# Add forecast data
forecast_part <- forecast_data[forecast_data$Quarter_Year %in% horizon_dates,]
p <- p %>% add_trace(
  data = forecast_part,
  x = ~Quarter_Year,
  y = ~Forecasted,
  type = 'scatter',
  mode = 'lines',
  name = "Projection",
  line = list(color = "#fdbf11", dash = "dash"),
  hoverinfo = "text",
  text = ~paste0(
    format(Quarter_Year, "%Y Q"), ceiling(as.numeric(format(Quarter_Year, "%m"))/3),
    "<br>Projected: ", formatC(Forecasted, format="d", big.mark=",")
  )
)

# Get all years for x-axis ticks
all_quarters <- unique(forecast_data$Quarter_Year)
all_years <- unique(format(all_quarters, "%Y"))

# Select every other year for x-axis ticks, using Q1 of each year
tick_positions <- c()
tick_labels <- c()

for (i in seq(1, length(all_years), by = 2)) {
  year <- all_years[i]
  year_quarters <- all_quarters[format(all_quarters, "%Y") == year]
  
  if (length(year_quarters) > 0) {
    # Find Q1 or closest quarter
    months <- as.numeric(format(year_quarters, "%m"))
    q1_idx <- which.min(abs(months - 2))  # closest to February (middle of Q1)
    q1_date <- year_quarters[q1_idx]
    
    tick_positions <- c(tick_positions, q1_date)
    tick_labels <- c(tick_labels, paste0(year, " Q", ceiling(months[q1_idx]/3)))
  }
}

# Complete layout
p %>% layout(
  title = title,
  xaxis = list(
    title = "Year-Quarter",
    type = "date",
    tickmode = "array",
    tickvals = tick_positions,
    ticktext = tick_labels,
    tickangle = 0
  ),
  yaxis = list(
    title = "Prison Population", 
    tickformat = ","
  ),
  hovermode = "closest",
  showlegend = TRUE
)


```

### Column {width=30%}

```{r}
#| label: total-projection-table

# Get the correct table data for total population
table_data <- total_data$table_data

# Limit rows to selected horizon
table_data <- head(table_data, projection_horizon)

# Format table
table_data <- table_data %>%
  mutate(
    Quarter_Year = as.character(Quarter_Year),
    Forecasted_hybrid = round(Forecasted_hybrid)
#     Lower_CI = round(Lower_CI),
#     Upper_CI = round(Upper_CI)
  ) %>%
  select(
    Quarter = Quarter_Year,
    Forecast = Forecasted_hybrid,
#     `Lower CI` = Lower_CI,
#     `Upper CI` = Upper_CI,
    `% Change` = Percent_Change
  )

datatable(
  table_data,
  options = list(
    pageLength = projection_horizon,
    dom = 't',
    ordering = FALSE
  ),
  rownames = FALSE,
  caption = "Total Population Projections"
) %>%
  formatStyle(
    '% Change',
    color = styleInterval(c(0), c('red', 'green')),
    fontWeight = 'bold'
  )
```

## Male Population

### Column {width=70%}

```{r}
#| label: male-projection-plot

# Get the appropriate forecast data for male population
male_data <- get_data_for_population("male")
forecast_data <- male_data$forecast_data
color_actual <- male_data$color
title <- paste(male_data$title, "Projection")

# Find the transition point between actual and forecast
actual_data <- forecast_data[!is.na(forecast_data$Actual),]
max_date <- max(actual_data$Quarter_Year)

# Get horizon dates for filtering
horizon_dates <- seq(max_date + 1, by = 1, length.out = projection_horizon)

# Create plot
p <- plot_ly(
  hoverlabel = list(namelength = -1)  # Don't truncate trace names in hover
)

# Add historical data
p <- p %>% add_trace(
  data = forecast_data[!is.na(forecast_data$Actual),],
  x = ~Quarter_Year,
  y = ~Actual,
  type = 'scatter',
  mode = 'lines',
  name = "Historical",
  line = list(color = color_actual),
  hoverinfo = "text",
  text = ~paste0(
    format(Quarter_Year, "%Y Q"), ceiling(as.numeric(format(Quarter_Year, "%m"))/3),
    "<br>Population: ", formatC(Actual, format="d", big.mark=",")
  )
)

# Add forecast data
forecast_part <- forecast_data[forecast_data$Quarter_Year %in% horizon_dates,]
p <- p %>% add_trace(
  data = forecast_part,
  x = ~Quarter_Year,
  y = ~Forecasted,
  type = 'scatter',
  mode = 'lines',
  name = "Projection",
  line = list(color = "#fdbf11", dash = "dash"),
  hoverinfo = "text",
  text = ~paste0(
    format(Quarter_Year, "%Y Q"), ceiling(as.numeric(format(Quarter_Year, "%m"))/3),
    "<br>Projected: ", formatC(Forecasted, format="d", big.mark=",")
  )
)

# Get all years for x-axis ticks
all_quarters <- unique(forecast_data$Quarter_Year)
all_years <- unique(format(all_quarters, "%Y"))

# Select every other year for x-axis ticks, using Q1 of each year
tick_positions <- c()
tick_labels <- c()

for (i in seq(1, length(all_years), by = 2)) {
  year <- all_years[i]
  year_quarters <- all_quarters[format(all_quarters, "%Y") == year]
  
  if (length(year_quarters) > 0) {
    # Find Q1 or closest quarter
    months <- as.numeric(format(year_quarters, "%m"))
    q1_idx <- which.min(abs(months - 2))  # closest to February (middle of Q1)
    q1_date <- year_quarters[q1_idx]
    
    tick_positions <- c(tick_positions, q1_date)
    tick_labels <- c(tick_labels, paste0(year, " Q", ceiling(months[q1_idx]/3)))
  }
}

# Complete layout
p %>% layout(
  title = title,
  xaxis = list(
    title = "Year-Quarter",
    type = "date",
    tickmode = "array",
    tickvals = tick_positions,
    ticktext = tick_labels,
    tickangle = 0
  ),
  yaxis = list(
    title = "Prison Population", 
    tickformat = ","
  ),
  hovermode = "closest",
  showlegend = TRUE
)

```

### Column {width=30%}

```{r}
#| label: male-projection-table

# Get the correct table data for male population
table_data <- male_data$table_data

# Limit rows to selected horizon
table_data <- head(table_data, projection_horizon)

# Format table
table_data <- table_data %>%
  mutate(
    Quarter_Year = as.character(Quarter_Year),
    Forecasted_hybrid = round(Forecasted_hybrid)
#     Lower_CI = round(Lower_CI),
#     Upper_CI = round(Upper_CI)
  ) %>%
  select(
    Quarter = Quarter_Year,
    Forecast = Forecasted_hybrid,
#     `Lower CI` = Lower_CI,
#     `Upper CI` = Upper_CI,
    `% Change` = Percent_Change
  )

datatable(
  table_data,
  options = list(
    pageLength = projection_horizon,
    dom = 't',
    ordering = FALSE
  ),
  rownames = FALSE,
  caption = "Male Population Projections"
) %>%
  formatStyle(
    '% Change',
    color = styleInterval(c(0), c('red', 'green')),
    fontWeight = 'bold'
  )
```

## Female Population

### Column {width=70%}

```{r}
#| label: female-projection-plot

# Get the appropriate forecast data for female population
female_data <- get_data_for_population("female")
forecast_data <- female_data$forecast_data
color_actual <- female_data$color
title <- paste(female_data$title, "Projection")

# Find the transition point between actual and forecast
actual_data <- forecast_data[!is.na(forecast_data$Actual),]
max_date <- max(actual_data$Quarter_Year)

# Get horizon dates for filtering
horizon_dates <- seq(max_date + 1, by = 1, length.out = projection_horizon)

# Create plot
p <- plot_ly(
  hoverlabel = list(namelength = -1)  # Don't truncate trace names in hover
)

# Add historical data
p <- p %>% add_trace(
  data = forecast_data[!is.na(forecast_data$Actual),],
  x = ~Quarter_Year,
  y = ~Actual,
  type = 'scatter',
  mode = 'lines',
  name = "Historical",
  line = list(color = color_actual),
  hoverinfo = "text",
  text = ~paste0(
    format(Quarter_Year, "%Y Q"), ceiling(as.numeric(format(Quarter_Year, "%m"))/3),
    "<br>Population: ", formatC(Actual, format="d", big.mark=",")
  )
)

# Add forecast data
forecast_part <- forecast_data[forecast_data$Quarter_Year %in% horizon_dates,]
p <- p %>% add_trace(
  data = forecast_part,
  x = ~Quarter_Year,
  y = ~Forecasted,
  type = 'scatter',
  mode = 'lines',
  name = "Projection",
  line = list(color = "#fdbf11", dash = "dash"),
  hoverinfo = "text",
  text = ~paste0(
    format(Quarter_Year, "%Y Q"), ceiling(as.numeric(format(Quarter_Year, "%m"))/3),
    "<br>Projected: ", formatC(Forecasted, format="d", big.mark=",")
  )
)

# Get all years for x-axis ticks
all_quarters <- unique(forecast_data$Quarter_Year)
all_years <- unique(format(all_quarters, "%Y"))

# Select every other year for x-axis ticks, using Q1 of each year
tick_positions <- c()
tick_labels <- c()

for (i in seq(1, length(all_years), by = 2)) {
  year <- all_years[i]
  year_quarters <- all_quarters[format(all_quarters, "%Y") == year]
  
  if (length(year_quarters) > 0) {
    # Find Q1 or closest quarter
    months <- as.numeric(format(year_quarters, "%m"))
    q1_idx <- which.min(abs(months - 2))  # closest to February (middle of Q1)
    q1_date <- year_quarters[q1_idx]
    
    tick_positions <- c(tick_positions, q1_date)
    tick_labels <- c(tick_labels, paste0(year, " Q", ceiling(months[q1_idx]/3)))
  }
}

# Complete layout
p %>% layout(
  title = title,
  xaxis = list(
    title = "Year-Quarter",
    type = "date",
    tickmode = "array",
    tickvals = tick_positions,
    ticktext = tick_labels,
    tickangle = 0
  ),
  yaxis = list(
    title = "Prison Population", 
    tickformat = ","
  ),
  hovermode = "closest",
  showlegend = TRUE
)
```

### Column {width=30%}

```{r}
#| label: female-projection-table

# Get the correct table data for female population
table_data <- female_data$table_data

# Limit rows to selected horizon
table_data <- head(table_data, projection_horizon)

# Format table
table_data <- table_data %>%
  mutate(
    Quarter_Year = as.character(Quarter_Year),
    Forecasted_hybrid = round(Forecasted_hybrid)
#    Lower_CI = round(Lower_CI),
#    Upper_CI = round(Upper_CI)
  ) %>%
  select(
    Quarter = Quarter_Year,
    Forecast = Forecasted_hybrid,
#     `Lower CI` = Lower_CI,
#     `Upper CI` = Upper_CI,
    `% Change` = Percent_Change
  )

datatable(
  table_data,
  options = list(
    pageLength = projection_horizon,
    dom = 't',
    ordering = FALSE
  ),
  rownames = FALSE,
  caption = "Female Population Projections"
) %>%
  formatStyle(
    '% Change',
    color = styleInterval(c(0), c('red', 'green')),
    fontWeight = 'bold'
  )
```

## Row {height="20%"}

::: callout-note
For detailed `confidence interval` data for these projections, see the [methodology](/#methodology).
:::


# ↕️️ Prison Drivers

::: panel-tabset
### Gender & Offense Type
```{r, warning=FALSE}
# Use the pre-processed driver data that was loaded
p <- df_driver_1 |> 
  filter(!is.na(gender), !is.na(top_ofnse_typ_desc)) |>
  mutate(Gender = as.character(gender)) |>
  ggplot(aes(x = Quarter_Year, y = Count,
             group = Gender, colour=Gender,
             text = paste("Quarter_Year:", format(Quarter_Year, "%Y Q%q"),
                          "<br>Count:", round(Count, 2),
                          "<br>Gender:", Gender,
                          "<br>Offense:", top_ofnse_typ_desc))) +
  stat_summary(fun = sum, geom = "line") +
  labs(title = "Utah prison population (by offense type & gender)",
       y = "Number of prisoners") +
  facet_wrap(~ as.character(top_ofnse_typ_desc),
             nrow = 4, scales = "free_y") +
  theme(axis.text.x = element_text(angle = 90)) +
  labs(y = "Number of prisoners",
       x = "Year/Quarter",
       title = "Utah Prison Population (by gender and offense type): 2010:Q1-Present") + 
  scale_color_manual(values = c("#00A7C4", "#DA9100", "grey"), 
                     labels = c("Female", "Male", "Unknown")) + 
  guides(color = guide_legend(title = "Gender")) + 
  scale_x_yearquarter(breaks = seq(yearquarter("2010 Q1"), yearquarter("2025 Q4"), by = 8),
                     labels = ~format(.x, "%Y Q%q")) + 
  theme(axis.text.x = element_text(angle = 90, hjust = 1))

p <- p + geom_vline(xintercept = yearquarter("2020 Q1"), linetype = "dashed", color = "grey")  # Add vertical line

# Convert to plotly with custom hover template
ggplotly(p, tooltip = "text")

```

### Gender & Admission Type
```{r}
# plot by admin_type type
p = df_driver_2 |>
  filter(!is_aggregated(gender), !is_aggregated(admin_type)) |>
  mutate(Gender = as.character(gender)) |>
  ggplot(aes(x = Quarter_Year, y = Count,
             group = Gender, colour=Gender,
             text = paste("Quarter_Year:", format(Quarter_Year, "%Y Q%q"),
                          "<br>Count:", round(Count, 2),
                          "<br>Gender:", Gender,
                          "<br>Admission Type:", admin_type))) +
  stat_summary(fun = sum, geom = "line") +
  labs(title = "Utah prison population (by admission type & gender)",
       y = "Number of prisoners") +
  facet_wrap(~ as.character(admin_type),
             nrow = 4, scales = "free_y") +
  theme(axis.text.x = element_text(angle = 90)) +
  labs(y = "Number of prisoners",
       x = "Year/Quarter",
       title = "Utah Prison Population (by gender and admission type): 2010:Q1-Present") + 
  scale_color_manual(values = c("#00A7C4", "#DA9100", "grey"), 
                     labels = c("Female", "Male", "Unknown")) + 
  guides(color = guide_legend(title = "Gender")) + 
  scale_x_yearquarter(breaks = seq(yearquarter("2010 Q1"), yearquarter("2025 Q4"), by = 8),
                      labels = ~format(.x, "%Y Q%q")) + 
  theme(axis.text.x = element_text(angle = 90, hjust = 1))

p = p + geom_vline(xintercept = yearquarter("2020 Q1"), linetype = "dashed", color = "grey")  # Add vertical line

# Convert to plotly with custom hover template
ggplotly(p, tooltip = "text")

```

### Offense Type & Admission
```{r}
# plot by admin/offense type
p = df_driver_3 |> 
  filter(!is_aggregated(admin_type), !is_aggregated(top_ofnse_typ_desc)) |>
  mutate(Admin = as.character(admin_type)) |>
  ggplot(aes(x = Quarter_Year, y = Count,
             group = Admin, colour=Admin,
             text = paste("Quarter_Year:", format(Quarter_Year, "%Y Q%q"),
                          "<br>Count:", round(Count, 2),
                          "<br>Admission Type:", Admin,
                          "<br>Offense Type:", top_ofnse_typ_desc))) +
  stat_summary(fun = sum, geom = "line") +
  labs(title = "Utah prison population (by offense type & admission)",
       y = "Number of prisoners") +
  facet_wrap(~ as.character(top_ofnse_typ_desc),
             nrow = 4, scales = "free_y") +
  theme(axis.text.x = element_text(angle = 90)) +
  labs(y = "Number of prisoners",
       x = "Year/Quarter",
       title = "Utah Prison Population (by admission type): 2010:Q1-Present") + 
  scale_color_manual(values = c("#9925be", "#be4d25", "grey")) + 
  guides(color = guide_legend(title = "Admission Type")) + 
  scale_x_yearquarter(breaks = seq(yearquarter("2010 Q1"), yearquarter("2025 Q4"), by = 8),
                      labels = ~format(.x, "%Y Q%q")) + 
  theme(axis.text.x = element_text(angle = 90, hjust = 1))

p = p + geom_vline(xintercept = yearquarter("2020 Q1"), linetype = "dashed", color = "grey")  # Add vertical line

# Convert to plotly with custom hover template
ggplotly(p, tooltip = "text")

```
::: 

# ⚠️ Capacity {.tabset}

## Total Population

```{r}
#| label: total-capacity-analysis

# Get the appropriate forecast data for total population
total_data <- get_data_for_population("total")
forecast_data <- total_data$forecast_data
color_actual <- total_data$color
title <- paste(total_data$title, "vs. Capacity")

# Get the capacity values directly from the function return values
absolute_max_capacity <- total_data$capacity_absolute_max
emergency_capacity <- total_data$capacity_emergency
operational_capacity <- total_data$capacity_operational

# Find the transition point between actual and forecast
actual_data <- forecast_data[!is.na(forecast_data$Actual),]
max_date <- max(actual_data$Quarter_Year)

# Get horizon dates for filtering
horizon_dates <- seq(max_date + 1, by = 1, length.out = projection_horizon)

# Create capacity dataframe with all three capacity types
capacity_df <- data.frame(
  Quarter_Year = rep(unique(forecast_data$Quarter_Year), 3),
  Capacity = c(
    rep(absolute_max_capacity, length(unique(forecast_data$Quarter_Year))),
    rep(emergency_capacity, length(unique(forecast_data$Quarter_Year))),
    rep(operational_capacity, length(unique(forecast_data$Quarter_Year)))
  ),
  Type = c(
    rep("Absolute Maximum Capacity", length(unique(forecast_data$Quarter_Year))),
    rep("Emergency Capacity", length(unique(forecast_data$Quarter_Year))),
    rep("Operational Capacity", length(unique(forecast_data$Quarter_Year)))
  )
)

# Calculate when each capacity is reached
capacity_reached <- list(
  absolute_max = NA,
  emergency = NA,
  operational = NA
)

for (i in 1:length(horizon_dates)) {
  date <- horizon_dates[i]
  projected_value <- forecast_data$Forecasted[forecast_data$Quarter_Year == date]
  
  if (!is.na(projected_value)) {
    if (is.na(capacity_reached$absolute_max) && projected_value > absolute_max_capacity) {
      capacity_reached$absolute_max <- date
    }
    if (is.na(capacity_reached$emergency) && projected_value > emergency_capacity) {
      capacity_reached$emergency <- date
    }
    if (is.na(capacity_reached$operational) && projected_value > operational_capacity) {
      capacity_reached$operational <- date
    }
  }
}


# Create plot
p <- plot_ly(
  hoverlabel = list(namelength = -1)  # Don't truncate trace names in hover
) %>%
  # Historical data
  add_trace(
    data = forecast_data[!is.na(forecast_data$Actual),],
    x = ~Quarter_Year,
    y = ~Actual,
    type = 'scatter',
    mode = 'lines',
    name = "Historical",
    line = list(color = color_actual),
    hoverinfo = "text",
    text = ~paste0(
      format(Quarter_Year, "%Y Q"), ceiling(as.numeric(format(Quarter_Year, "%m"))/3),
      "<br>Population: ", formatC(Actual, format="d", big.mark=",")
    )
  )

# Add forecasted data for the selected horizon
forecast_part <- forecast_data[forecast_data$Quarter_Year %in% horizon_dates,]
p <- p %>%
  add_trace(
    data = forecast_part,
    x = ~Quarter_Year,
    y = ~Forecasted,
    type = 'scatter',
    mode = 'lines',
    name = "Projection",
    line = list(color = "#fdbf11", dash = "dash"),
    hoverinfo = "text",
    text = ~paste0(
      format(Quarter_Year, "%Y Q"), ceiling(as.numeric(format(Quarter_Year, "%m"))/3),
      "<br>Projected: ", formatC(Forecasted, format="d", big.mark=",")
    )
  )

# Add capacity lines
p <- p %>%
  # Absolute Maximum Capacity (red)
  add_trace(
    data = capacity_df[capacity_df$Type == "Absolute Maximum Capacity",],
    x = ~Quarter_Year,
    y = ~Capacity,
    type = 'scatter',
    mode = 'lines',
    name = "Absolute Maximum Capacity",
    line = list(color = "red", dash = "dot", width = 2),
    hoverinfo = "text",
    text = ~paste0(
      format(Quarter_Year, "%Y Q"), ceiling(as.numeric(format(Quarter_Year, "%m"))/3),
      "<br>Absolute Maximum Capacity: ", formatC(Capacity, format="d", big.mark=",")
    )
  ) %>%
  # Emergency Capacity (orange)
  add_trace(
    data = capacity_df[capacity_df$Type == "Emergency Capacity",],
    x = ~Quarter_Year,
    y = ~Capacity,
    type = 'scatter',
    mode = 'lines',
    name = "Emergency Capacity",
    line = list(color = "brown", dash = "dot", width = 2),
    hoverinfo = "text",
    text = ~paste0(
      format(Quarter_Year, "%Y Q"), ceiling(as.numeric(format(Quarter_Year, "%m"))/3),
      "<br>Emergency Capacity: ", formatC(Capacity, format="d", big.mark=",")
    )
  ) %>%
  # Operational Capacity (purple)
  add_trace(
    data = capacity_df[capacity_df$Type == "Operational Capacity",],
    x = ~Quarter_Year,
    y = ~Capacity,
    type = 'scatter',
    mode = 'lines',
    name = "Operational Capacity",
    line = list(color = "green", dash = "dot", width = 2),
    hoverinfo = "text",
    text = ~paste0(
      format(Quarter_Year, "%Y Q"), ceiling(as.numeric(format(Quarter_Year, "%m"))/3),
      "<br>Operational Capacity: ", formatC(Capacity, format="d", big.mark=",")
    )
  )

# Add capacity reached annotations if applicable
for (capacity_type in names(capacity_reached)) {
  date <- capacity_reached[[capacity_type]]
  if (!is.na(date)) {
    projected_value <- forecast_data$Forecasted[forecast_data$Quarter_Year == date]
    
    # Format quarter
    year <- format(date, "%Y")
    quarter <- paste0("Q", ceiling(as.numeric(format(date, "%m"))/3))
    quarter_label <- paste0(year, " ", quarter)
    
    capacity_name <- switch(capacity_type,
                         "absolute_max" = "Absolute Maximum Capacity",
                         "emergency" = "Emergency Capacity",
                         "operational" = "Operational Capacity")
    
    annotation_color <- switch(capacity_type,
                            "absolute_max" = "red",
                            "emergency" = "brown",
                            "operational" = "green")
    
    p <- p %>%
      add_annotations(
        x = date,
        y = projected_value,
        text = paste(capacity_name, "reached:", quarter_label),
        showarrow = TRUE,
        arrowhead = 1,
        arrowsize = 1,
        arrowwidth = 2,
        arrowcolor = annotation_color,
        ax = 0, 
        ay = -40 * match(capacity_type, names(capacity_reached)), # Stagger annotations
        font = list(color = annotation_color)
      )
  }
}

# Get years for x-axis ticks
all_quarters <- unique(forecast_data$Quarter_Year)
q1_dates <- c()
q1_labels <- c()

# Find Q1 for each year
all_years <- unique(format(all_quarters, "%Y"))
for (year in all_years) {
  year_quarters <- all_quarters[format(all_quarters, "%Y") == year]
  months <- as.numeric(format(year_quarters, "%m"))
  q1_idx <- which.min(abs(months - 2)) # Closest to Q1 (Feb is middle of Q1)
  
  if (length(q1_idx) > 0) {
    q1_date <- year_quarters[q1_idx]
    q1_labels <- c(q1_labels, paste0(year, " Q", ceiling(months[q1_idx]/3)))
  }
}

# Complete the plot with formatting
p %>% layout(
  title = title,
  xaxis = list(
    title = "Year-Quarter",
    type = "date",
    tickmode = "array",
    tickvals = all_quarters[seq(1, length(all_quarters), by = 4)],  # Show every year
    ticktext = format(all_quarters[seq(1, length(all_quarters), by = 4)], "%Y Q%q"),
    tickangle = 0
  ),
  yaxis = list(
    title = "Prison Population", 
    tickformat = ","
  ),
  hovermode = "closest",
  showlegend = TRUE
)


```

## Male Population

```{r}
#| label: male-capacity-analysis

# Get the appropriate forecast data for male population
male_data <- get_data_for_population("male")
forecast_data <- male_data$forecast_data
color_actual <- male_data$color
title <- paste(male_data$title, "vs. Capacity")

# Get the capacity values directly from the function return values
absolute_max_capacity <- 6588  # Male-specific absolute maximum capacity
emergency_capacity <- 6456     # Example value - adjust as needed
operational_capacity <- 6357   # Example value - adjust as needed

# Find the transition point between actual and forecast
actual_data <- forecast_data[!is.na(forecast_data$Actual),]
max_date <- max(actual_data$Quarter_Year)

# Get horizon dates for filtering
horizon_dates <- seq(max_date + 1, by = 1, length.out = projection_horizon)

# Create capacity dataframe with all three capacity types
capacity_df <- data.frame(
  Quarter_Year = rep(unique(forecast_data$Quarter_Year), 3),
  Capacity = c(
    rep(absolute_max_capacity, length(unique(forecast_data$Quarter_Year))),
    rep(emergency_capacity, length(unique(forecast_data$Quarter_Year))),
    rep(operational_capacity, length(unique(forecast_data$Quarter_Year)))
  ),
  Type = c(
    rep("Absolute Maximum Capacity", length(unique(forecast_data$Quarter_Year))),
    rep("Emergency Capacity", length(unique(forecast_data$Quarter_Year))),
    rep("Operational Capacity", length(unique(forecast_data$Quarter_Year)))
  )
)

# Calculate when each capacity is reached
capacity_reached <- list(
  absolute_max = NA,
  emergency = NA,
  operational = NA
)

for (i in 1:length(horizon_dates)) {
  date <- horizon_dates[i]
  projected_value <- forecast_data$Forecasted[forecast_data$Quarter_Year == date]
  
  if (!is.na(projected_value)) {
    if (is.na(capacity_reached$absolute_max) && projected_value > absolute_max_capacity) {
      capacity_reached$absolute_max <- date
    }
    if (is.na(capacity_reached$emergency) && projected_value > emergency_capacity) {
      capacity_reached$emergency <- date
    }
    if (is.na(capacity_reached$operational) && projected_value > operational_capacity) {
      capacity_reached$operational <- date
    }
  }
}

# Create plot
p <- plot_ly(
  hoverlabel = list(namelength = -1)  # Don't truncate trace names in hover
) %>%
  # Historical data
  add_trace(
    data = forecast_data[!is.na(forecast_data$Actual),],
    x = ~Quarter_Year,
    y = ~Actual,
    type = 'scatter',
    mode = 'lines',
    name = "Historical",
    line = list(color = color_actual),
    hoverinfo = "text",
    text = ~paste0(
      format(Quarter_Year, "%Y Q"), ceiling(as.numeric(format(Quarter_Year, "%m"))/3),
      "<br>Population: ", formatC(Actual, format="d", big.mark=",")
    )
  )

# Add forecasted data for the selected horizon
forecast_part <- forecast_data[forecast_data$Quarter_Year %in% horizon_dates,]
p <- p %>%
  add_trace(
    data = forecast_part,
    x = ~Quarter_Year,
    y = ~Forecasted,
    type = 'scatter',
    mode = 'lines',
    name = "Projection",
    line = list(color = "#fdbf11", dash = "dash"),
    hoverinfo = "text",
    text = ~paste0(
      format(Quarter_Year, "%Y Q"), ceiling(as.numeric(format(Quarter_Year, "%m"))/3),
      "<br>Projected: ", formatC(Forecasted, format="d", big.mark=",")
    )
  )

# Add capacity lines
p <- p %>%
  # Absolute Maximum Capacity (red)
  add_trace(
    data = capacity_df[capacity_df$Type == "Absolute Maximum Capacity",],
    x = ~Quarter_Year,
    y = ~Capacity,
    type = 'scatter',
    mode = 'lines',
    name = "Absolute Maximum Capacity",
    line = list(color = "red", dash = "dot", width = 2),
    hoverinfo = "text",
    text = ~paste0(
      format(Quarter_Year, "%Y Q"), ceiling(as.numeric(format(Quarter_Year, "%m"))/3),
      "<br>Absolute Maximum Capacity: ", formatC(Capacity, format="d", big.mark=",")
    )
  ) %>%
  # Emergency Capacity (orange)
  add_trace(
    data = capacity_df[capacity_df$Type == "Emergency Capacity",],
    x = ~Quarter_Year,
    y = ~Capacity,
    type = 'scatter',
    mode = 'lines',
    name = "Emergency Capacity",
    line = list(color = "brown", dash = "dot", width = 2),
    hoverinfo = "text",
    text = ~paste0(
      format(Quarter_Year, "%Y Q"), ceiling(as.numeric(format(Quarter_Year, "%m"))/3),
      "<br>Emergency Capacity: ", formatC(Capacity, format="d", big.mark=",")
    )
  ) %>%
  # Operational Capacity (purple)
  add_trace(
    data = capacity_df[capacity_df$Type == "Operational Capacity",],
    x = ~Quarter_Year,
    y = ~Capacity,
    type = 'scatter',
    mode = 'lines',
    name = "Operational Capacity",
    line = list(color = "green", dash = "dot", width = 2),
    hoverinfo = "text",
    text = ~paste0(
      format(Quarter_Year, "%Y Q"), ceiling(as.numeric(format(Quarter_Year, "%m"))/3),
      "<br>Operational Capacity: ", formatC(Capacity, format="d", big.mark=",")
    )
  )

# Add capacity reached annotations if applicable
for (capacity_type in names(capacity_reached)) {
  date <- capacity_reached[[capacity_type]]
  if (!is.na(date)) {
    projected_value <- forecast_data$Forecasted[forecast_data$Quarter_Year == date]
    
    # Format quarter
    year <- format(date, "%Y")
    quarter <- paste0("Q", ceiling(as.numeric(format(date, "%m"))/3))
    quarter_label <- paste0(year, " ", quarter)
    
    capacity_name <- switch(capacity_type,
                         "absolute_max" = "Absolute Maximum Capacity",
                         "emergency" = "Emergency Capacity",
                         "operational" = "Operational Capacity")
    
    annotation_color <- switch(capacity_type,
                            "absolute_max" = "red",
                            "emergency" = "brown",
                            "operational" = "green")
    
    p <- p %>%
      add_annotations(
        x = date,
        y = projected_value,
        text = paste(capacity_name, "reached:", quarter_label),
        showarrow = TRUE,
        arrowhead = 1,
        arrowsize = 1,
        arrowwidth = 2,
        arrowcolor = annotation_color,
        ax = 0, 
        ay = -40 * match(capacity_type, names(capacity_reached)), # Stagger annotations
        font = list(color = annotation_color)
      )
  }
}

# Get years for x-axis ticks
all_quarters <- unique(forecast_data$Quarter_Year)
q1_dates <- c()
q1_labels <- c()

# Find Q1 for each year
all_years <- unique(format(all_quarters, "%Y"))
for (year in all_years) {
  year_quarters <- all_quarters[format(all_quarters, "%Y") == year]
  months <- as.numeric(format(year_quarters, "%m"))
  q1_idx <- which.min(abs(months - 2)) # Closest to Q1 (Feb is middle of Q1)
  
  if (length(q1_idx) > 0) {
    q1_date <- year_quarters[q1_idx]
    q1_labels <- c(q1_labels, paste0(year, " Q", ceiling(months[q1_idx]/3)))
  }
}

# Complete the plot with formatting
p %>% layout(
  title = title,
  xaxis = list(
    title = "Year-Quarter",
    type = "date",
    tickmode = "array",
    tickvals = all_quarters[seq(1, length(all_quarters), by = 4)],  # Show every year
    ticktext = format(all_quarters[seq(1, length(all_quarters), by = 4)], "%Y Q%q"),
    tickangle = 0
  ),
  yaxis = list(
    title = "Prison Population", 
    tickformat = ","
  ),
  hovermode = "closest",
  showlegend = TRUE
)


```

## Female Population

```{r}
#| label: female-capacity-analysis

# Get the appropriate forecast data for female population
female_data <- get_data_for_population("female")
forecast_data <- female_data$forecast_data
color_actual <- female_data$color
title <- paste(female_data$title, "vs. Capacity")

# Get the capacity values directly from the function return values
absolute_max_capacity <- 632   # Female-specific absolute maximum capacity
emergency_capacity <- 619      # Example value - adjust as needed
operational_capacity <- 610   # Example value - adjust as needed

# Find the transition point between actual and forecast
actual_data <- forecast_data[!is.na(forecast_data$Actual),]
max_date <- max(actual_data$Quarter_Year)

# Get horizon dates for filtering
horizon_dates <- seq(max_date + 1, by = 1, length.out = projection_horizon)

# Create capacity dataframe with all three capacity types
capacity_df <- data.frame(
  Quarter_Year = rep(unique(forecast_data$Quarter_Year), 3),
  Capacity = c(
    rep(absolute_max_capacity, length(unique(forecast_data$Quarter_Year))),
    rep(emergency_capacity, length(unique(forecast_data$Quarter_Year))),
    rep(operational_capacity, length(unique(forecast_data$Quarter_Year)))
  ),
  Type = c(
    rep("Absolute Maximum Capacity", length(unique(forecast_data$Quarter_Year))),
    rep("Emergency Capacity", length(unique(forecast_data$Quarter_Year))),
    rep("Operational Capacity", length(unique(forecast_data$Quarter_Year)))
  )
)

# Calculate when each capacity is reached
capacity_reached <- list(
  absolute_max = NA,
  emergency = NA,
  operational = NA
)

for (i in 1:length(horizon_dates)) {
  date <- horizon_dates[i]
  projected_value <- forecast_data$Forecasted[forecast_data$Quarter_Year == date]
  
  if (!is.na(projected_value)) {
    if (is.na(capacity_reached$absolute_max) && projected_value > absolute_max_capacity) {
      capacity_reached$absolute_max <- date
    }
    if (is.na(capacity_reached$emergency) && projected_value > emergency_capacity) {
      capacity_reached$emergency <- date
    }
    if (is.na(capacity_reached$operational) && projected_value > operational_capacity) {
      capacity_reached$operational <- date
    }
  }
}

# Create plot
p <- plot_ly(
  hoverlabel = list(namelength = -1)  # Don't truncate trace names in hover
) %>%
  # Historical data
  add_trace(
    data = forecast_data[!is.na(forecast_data$Actual),],
    x = ~Quarter_Year,
    y = ~Actual,
    type = 'scatter',
    mode = 'lines',
    name = "Historical",
    line = list(color = color_actual),
    hoverinfo = "text",
    text = ~paste0(
      format(Quarter_Year, "%Y Q"), ceiling(as.numeric(format(Quarter_Year, "%m"))/3),
      "<br>Population: ", formatC(Actual, format="d", big.mark=",")
    )
  )

# Add forecasted data for the selected horizon
forecast_part <- forecast_data[forecast_data$Quarter_Year %in% horizon_dates,]
p <- p %>%
  add_trace(
    data = forecast_part,
    x = ~Quarter_Year,
    y = ~Forecasted,
    type = 'scatter',
    mode = 'lines',
    name = "Projection",
    line = list(color = "#fdbf11", dash = "dash"),
    hoverinfo = "text",
    text = ~paste0(
      format(Quarter_Year, "%Y Q"), ceiling(as.numeric(format(Quarter_Year, "%m"))/3),
      "<br>Projected: ", formatC(Forecasted, format="d", big.mark=",")
    )
  )

# Add capacity lines
p <- p %>%
  # Absolute Maximum Capacity (red)
  add_trace(
    data = capacity_df[capacity_df$Type == "Absolute Maximum Capacity",],
    x = ~Quarter_Year,
    y = ~Capacity,
    type = 'scatter',
    mode = 'lines',
    name = "Absolute Maximum Capacity",
    line = list(color = "red", dash = "dot", width = 2),
    hoverinfo = "text",
    text = ~paste0(
      format(Quarter_Year, "%Y Q"), ceiling(as.numeric(format(Quarter_Year, "%m"))/3),
      "<br>Absolute Maximum Capacity: ", formatC(Capacity, format="d", big.mark=",")
    )
  ) %>%
  # Emergency Capacity (orange)
  add_trace(
    data = capacity_df[capacity_df$Type == "Emergency Capacity",],
    x = ~Quarter_Year,
    y = ~Capacity,
    type = 'scatter',
    mode = 'lines',
    name = "Emergency Capacity",
    line = list(color = "brown", dash = "dot", width = 2),
    hoverinfo = "text",
    text = ~paste0(
      format(Quarter_Year, "%Y Q"), ceiling(as.numeric(format(Quarter_Year, "%m"))/3),
      "<br>Emergency Capacity: ", formatC(Capacity, format="d", big.mark=",")
    )
  ) %>%
  # Operational Capacity (purple)
  add_trace(
    data = capacity_df[capacity_df$Type == "Operational Capacity",],
    x = ~Quarter_Year,
    y = ~Capacity,
    type = 'scatter',
    mode = 'lines',
    name = "Operational Capacity",
    line = list(color = "green", dash = "dot", width = 2),
    hoverinfo = "text",
    text = ~paste0(
      format(Quarter_Year, "%Y Q"), ceiling(as.numeric(format(Quarter_Year, "%m"))/3),
      "<br>Operational Capacity: ", formatC(Capacity, format="d", big.mark=",")
    )
  )

# Add capacity reached annotations if applicable
for (capacity_type in names(capacity_reached)) {
  date <- capacity_reached[[capacity_type]]
  if (!is.na(date)) {
    projected_value <- forecast_data$Forecasted[forecast_data$Quarter_Year == date]
    
    # Format quarter
    year <- format(date, "%Y")
    quarter <- paste0("Q", ceiling(as.numeric(format(date, "%m"))/3))
    quarter_label <- paste0(year, " ", quarter)
    
    capacity_name <- switch(capacity_type,
                         "absolute_max" = "Absolute Maximum Capacity",
                         "emergency" = "Emergency Capacity",
                         "operational" = "Operational Capacity")
    
    annotation_color <- switch(capacity_type,
                            "absolute_max" = "red",
                            "emergency" = "brown",
                            "operational" = "green")
    
    p <- p %>%
      add_annotations(
        x = date,
        y = projected_value,
        text = paste(capacity_name, "reached:", quarter_label),
        showarrow = TRUE,
        arrowhead = 1,
        arrowsize = 1,
        arrowwidth = 2,
        arrowcolor = annotation_color,
        ax = 0, 
        ay = -40 * match(capacity_type, names(capacity_reached)), # Stagger annotations
        font = list(color = annotation_color)
      )
  }
}

# Get years for x-axis ticks
all_quarters <- unique(forecast_data$Quarter_Year)
q1_dates <- c()
q1_labels <- c()

# Find Q1 for each year
all_years <- unique(format(all_quarters, "%Y"))
for (year in all_years) {
  year_quarters <- all_quarters[format(all_quarters, "%Y") == year]
  months <- as.numeric(format(year_quarters, "%m"))
  q1_idx <- which.min(abs(months - 2)) # Closest to Q1 (Feb is middle of Q1)
  
  if (length(q1_idx) > 0) {
    q1_date <- year_quarters[q1_idx]
    q1_labels <- c(q1_labels, paste0(year, " Q", ceiling(months[q1_idx]/3)))
  }
}

# Complete the plot with formatting
p %>% layout(
  title = title,
  xaxis = list(
    title = "Year-Quarter",
    type = "date",
    tickmode = "array",
    tickvals = all_quarters[seq(1, length(all_quarters), by = 4)],  # Show every year
    ticktext = format(all_quarters[seq(1, length(all_quarters), by = 4)], "%Y Q%q"),
    tickangle = 0
  ),
  yaxis = list(
    title = "Prison Population", 
    tickformat = ","
  ),
  hovermode = "closest",
  showlegend = TRUE
)

```


# 📝 Methodology

## Row

```{r}
#| label: pdf-embed
#| echo: false

HTML('
<div class="methodology-card">
  <h3>Prison Population Projection Methodology</h3>
</div>
')

htmltools::tags$iframe(
  src = "UPPP Methodology.pdf",
  width = "100%",
  height = "700px",
  style = "border: 1px solid #ddd;"
)


```

# 📋 Policy Briefs

## Row

```{r}
#| label: pdf-embed2
#| echo: false

HTML('
<div class="methodology-card">
  <h3>Prison Population Projection Policy Briefs</h3>
</div>
')

htmltools::tags$iframe(
  src = "UPPP Policy Brief 02.pdf",
  width = "100%",
  height = "700px",
  style = "border: 1px solid #ddd;"
)

```

