---
title: "Utah Prison Population Projections"
subtitle: "Historical Trends and Future Projections"
format: 
  dashboard:
    orientation: rows  
    logo: images/CCJJ_Logo_1.png
    theme: 
      - sandstone
      - custom.scss
    nav-buttons:
      - linkedin
      - twitter
      - icon: gitlab
        href: https://gitlab.com/
    scrolling: false
---

```{r}
#| label: setup
#| message: false
#| include: false

# Load required packages - streamlined to only what's needed for visualization
library(tidyverse)
library(plotly)
library(DT)
library(scales)
library(bslib)
library(bsicons)
library(lubridate)
library(kableExtra)
library(knitr)
library(readxl)
library(dplyr)
library(lubridate)
library(tsibble)
library(ggplot2)
library(plotly)
library(scales)
library(DT)
library(forecast)
library(htmltools)  # For tags$p function

# Set global options
options(scipen = 999)
theme_set(theme_minimal())

# Define parameters that would have been selected by the user
population_type <- "total"  # Options: "total", "male", "female"
projection_horizon <- 20    # Options: 8, 12, 20
scenario <- "baseline"      # Options: "baseline", "reform", "covid"
```

```{r}
#| label: load-preprocessed-data
#| message: false

# Load pre-processed data files
prison_data_clean <- read.csv("data/prison_data_clean.csv") %>%
  mutate(Quarter_Year = yearquarter(Quarter_Year))

# Load forecast dataframes
forecast_df <- read.csv("data/forecast_total.csv") %>%
  mutate(Quarter_Year = yearquarter(Quarter_Year))

forecast_df_m <- read.csv("data/forecast_male.csv") %>%
  mutate(Quarter_Year = yearquarter(Quarter_Year))

forecast_df_f <- read.csv("data/forecast_female.csv") %>%
  mutate(Quarter_Year = yearquarter(Quarter_Year))

# Load forecast tables
forecast_table_hybrid <- read.csv("data/forecast_table_total.csv") %>%
  mutate(Quarter_Year = yearquarter(Quarter_Year))

forecast_table_hybrid_m <- read.csv("data/forecast_table_male.csv") %>%
  mutate(Quarter_Year = yearquarter(Quarter_Year))

forecast_table_hybrid_f <- read.csv("data/forecast_table_female.csv") %>%
  mutate(Quarter_Year = yearquarter(Quarter_Year))

# Load summary metrics for value boxes
summary_metrics <- read.csv("data/summary_metrics.csv")

# Extract current metrics from summary data
current_total <- summary_metrics %>% filter(population_type == "total") %>% pull(current_pop)
current_male <- summary_metrics %>% filter(population_type == "male") %>% pull(current_pop)
current_female <- summary_metrics %>% filter(population_type == "female") %>% pull(current_pop)

projected_total <- summary_metrics %>% filter(population_type == "total") %>% pull(projected_pop)
projected_male <- summary_metrics %>% filter(population_type == "male") %>% pull(projected_pop)
projected_female <- summary_metrics %>% filter(population_type == "female") %>% pull(projected_pop)

pct_change_total <- summary_metrics %>% filter(population_type == "total") %>% pull(pct_change)
pct_change_male <- summary_metrics %>% filter(population_type == "male") %>% pull(pct_change)
pct_change_female <- summary_metrics %>% filter(population_type == "female") %>% pull(pct_change)

annual_change_total <- summary_metrics %>% filter(population_type == "total") %>% pull(annual_change)
annual_change_male <- summary_metrics %>% filter(population_type == "male") %>% pull(annual_change)
annual_change_female <- summary_metrics %>% filter(population_type == "female") %>% pull(annual_change)

# Read in Prison Admission Data
admin_rel_merged_df <- read.csv("data/admin_rel_merged_df.csv")

# Load capacity data
capacity_data <- read.csv("data/capacity_limits.csv")

# Load gender & offense type data for faceted plots
df_driver_1 <- read.csv("data/df_driver_1.csv") %>%
  mutate(Quarter_Year = yearquarter(Quarter_Year))

df_driver_2 <- read.csv("data/df_driver_2.csv") %>%
  mutate(Quarter_Year = yearquarter(Quarter_Year))

df_driver_3 <- read.csv("data/df_driver_3.csv") %>%
  mutate(Quarter_Year = yearquarter(Quarter_Year))

# Helper function to select appropriate data based on population type
get_data_for_population <- function(pop_type = population_type) {
  if (pop_type == "total") {
    list(
      forecast_data = forecast_df,
      table_data = forecast_table_hybrid,
      current_pop = current_total,
      projected_pop = projected_total,
      pct_change = pct_change_total,
      annual_change = annual_change_total,
      title = "Total Prison Population",
      color = "#00A7C4",
      capacity = capacity_data %>% filter(type == "total") %>% pull(capacity)
    )
  } else if (pop_type == "male") {
    list(
      forecast_data = forecast_df_m,
      table_data = forecast_table_hybrid_m,
      current_pop = current_male,
      projected_pop = projected_male,
      pct_change = pct_change_male,
      annual_change = annual_change_male,
      title = "Male Prison Population",
      color = "#0a4c6a",
      capacity = capacity_data %>% filter(type == "male") %>% pull(capacity)
    )
  } else {
    list(
      forecast_data = forecast_df_f,
      table_data = forecast_table_hybrid_f,
      current_pop = current_female,
      projected_pop = projected_female,
      pct_change = pct_change_female,
      annual_change = annual_change_female,
      title = "Female Prison Population",
      color = "#fc719e",
      capacity = capacity_data %>% filter(type == "female") %>% pull(capacity)
    )
  }
}

# Get data for the currently selected population type
selected_data <- get_data_for_population(population_type)
```

#  {.sidebar}

### Population Type:
Total Population

### Projection Horizon:
5 Years (20 Quarters)

### Policy Scenario:
Baseline

<hr>

This dashboard provides projections of Utah's prison population based on historical data. Use the tabs above to explore different aspects of the projections.

# 🏛️ Overview {orientation="rows"}

## Value Boxes {height="40%"}

```{r}
#| label: value-boxes

# Get data based on the selected population type
data <- get_data_for_population(population_type)

# Create layout with value boxes
layout_column_wrap(
  width = "250px",
  value_box(
    title = paste(str_to_title(gsub("Population", "", population_type)), "Current Population"),
    value = comma(round(data$current_pop)),
    showcase = bs_icon("people-fill"),
    theme = value_box_theme(bg = "#e6f2fd", fg = "#0B538E"),
    tags$p("2024 Q4")
  ),
  value_box(
    title = paste(str_to_title(gsub("Population", "", population_type)), "Projected Population"),
    value = comma(round(data$projected_pop)),
    showcase = bs_icon("graph-up"),
    theme = "teal",
    tags$p("5 years projection"),
    tags$p("2029 Q4")
  ),
  value_box(
    title = "Percent Change",
    value = paste0(data$pct_change, "%"),
    showcase = bs_icon("percent"),
    theme = "pink",
    tags$p("over 5 years")
  ),
  value_box(
    title = "Average Annual Change",
    value = paste0(data$annual_change, "%"),
    showcase = bs_icon("calendar-event"),
    theme = "warning",
    tags$p("annually")
  )
)
```

## Plot {height="60%"}

```{r}
#| label: admissions-releases-plot
#| title: Quarterly Net Difference Prison Admissions and Releases Over Time
#| height: 60%

# Plot difference
plot_ly(admin_rel_merged_df, x = ~as.character(Year_Quarter)) %>%
  add_lines(y = ~diff, name = "Net difference", line = list(color = "#00A7C4")) %>%
  layout(
    title = "Quarterly Net Difference: Prison Admissions and Releases",
    xaxis = list(
      title = "Year-Quarter",
      tickangle = 0,
      tickfont = list(size = 10)
    ),
    yaxis = list(title = "Count", range = c(0, NA)),  # Start y-axis at 0
    margin = list(b = 80)  # Add bottom margin for labels
  )
```

# 📈 Projections {.tabset}

## Total Population

### Row {height="50%"}

```{r}
#| label: total-projection-plot

# Get the appropriate forecast data for total population
total_data <- get_data_for_population("total")
forecast_data <- total_data$forecast_data
color_actual <- total_data$color
title <- "Actual (2010 Q1 - 2024 Q2) vs Forecasted Prison Population"

# Find the transition point between actual and forecast
actual_data <- forecast_data[!is.na(forecast_data$Actual),]
max_date <- max(actual_data$Quarter_Year)

# Get horizon dates for filtering
horizon_dates <- seq(max_date + 1, by = 1, length.out = projection_horizon)

# Create plot
p <- plot_ly()

# Add historical data
p <- p %>% add_lines(
  data = forecast_data[!is.na(forecast_data$Actual),],
  x = ~Quarter_Year,
  y = ~Actual,
  name = "Actual",
  line = list(color = color_actual)
)

# Add forecast data
forecast_part <- forecast_data[forecast_data$Quarter_Year %in% horizon_dates,]
p <- p %>% add_lines(
  data = forecast_part,
  x = ~Quarter_Year,
  y = ~Forecasted,
  name = "Forecasted",
  line = list(color = "#fdbf11", dash = "dash")
)

# Add capacity line
p <- p %>% add_lines(
  data = capacity_df,
  x = ~Quarter_Year,
  y = ~Capacity,
  name = "Capacity Limit",
  line = list(color = "red", dash = "dot", width = 2)
)

# Add capacity reached annotation if applicable
if (!is.na(capacity_reached)) {
  projected_value <- forecast_data$Forecasted[forecast_data$Quarter_Year == capacity_reached]
  p <- p %>% add_annotations(
    x = capacity_reached,
    y = projected_value,
    text = paste("Capacity reached:", format(capacity_reached, "%Y Q%q")),
    showarrow = TRUE,
    arrowhead = 1,
    arrowsize = 1,
    arrowwidth = 2,
    arrowcolor = "red",
    ax = 0,
    ay = -40,
    font = list(color = "red")
  )
}

# Complete the plot
p %>% layout(
  title = title,
  xaxis = list(
    title = "Quarter/Year",
    tickformat = "%Y Q%q",
    tickmode = "array",
    tickvals = c(
      yearquarter("2010 Q1"),
      yearquarter("2015 Q1"),
      yearquarter("2020 Q1"),
      yearquarter("2025 Q1"),
      yearquarter("2030 Q1")
    ),
    ticktext = c("2010 Q1", "2015 Q1", "2020 Q1", "2025 Q1", "2030 Q1")
  ),
  yaxis = list(title = "Prison Population", tickformat = ","),
  hovermode = "x unified",
  showlegend = TRUE
)
```


# 📝 Methodology

## Row

```{r}
HTML('
<div class="methodology-card">
  <h3>Prison Population Projection Methodology</h3>
  
  <h4>Hybrid Forecasting Approach</h4>
  <p>The projections in this dashboard use a hybrid modeling approach that combines:</p>
  
  <ol>
    <li><strong>Linear Regression with Structural Breaks:</strong> 
      The base model captures long-term trends and policy impacts, including:
      <ul>
        <li>Time trend component</li>
        <li>JRI (Justice Reinvestment Initiative) policy impact (2015)</li>
        <li>COVID-19 impact (2020)</li>
      </ul>
    </li>
    <li><strong>Exponential Smoothing (ETS):</strong> 
      Applied to the residuals of the linear model to capture short-term patterns, with alpha parameter set to 0.8 to emphasize recent patterns
    </li>
  </ol>
  
  <h4>Data Sources</h4>
  <p>The projections are based on historical prison population data by demographic groups, analyzed at the quarterly level to balance detail with forecasting stability.</p>
</div>
')
```

# Complete layout
```{r}
p %>% layout(
  title = title,
  xaxis = list(
    title = "Quarter/Year",
    tickformat = "%Y Q%q",
    tickmode = "array",
    tickvals = c(
      yearquarter("2010 Q1"),
      yearquarter("2015 Q1"),
      yearquarter("2020 Q1"),
      yearquarter("2025 Q1"),
      yearquarter("2030 Q1")
    ),
    ticktext = c("2010 Q1", "2015 Q1", "2020 Q1", "2025 Q1", "2030 Q1")
  ),
  yaxis = list(title = "Prison Population", tickformat = ","),
  hovermode = "x unified",
  showlegend = TRUE,
  legend = list(title = list(text = "Series"))
)
```

### Row {height="30%"}

```{r}
#| label: total-projection-table

# Get the correct table data for total population
table_data <- total_data$table_data

# Limit rows to selected horizon
table_data <- head(table_data, projection_horizon)

# Format table
table_data <- table_data %>%
  mutate(
    Quarter_Year = format(Quarter_Year, "%Y Q%q"),
    Forecasted_hybrid = round(Forecasted_hybrid, 0)
  ) %>%
  select(
    Quarter_Year,
    Forecasted_hybrid,
    Percent_Change
  )

datatable(
  table_data,
  options = list(
    pageLength = 5,
    dom = 't',
    ordering = FALSE
  ),
  colnames = c('Quarter Year' = 'Quarter_Year', 
               'Forecasted' = 'Forecasted_hybrid', 
               '% Change' = 'Percent_Change'),
  rownames = FALSE
) %>%
  formatStyle(
    'Percent_Change',
    color = styleInterval(c(0), c('red', 'green')),
    fontWeight = 'bold'
  ) %>%
  formatRound('Forecasted_hybrid', 0) %>%
  formatRound('Percent_Change', 1) %>%
  formatStyle('Forecasted_hybrid', fontWeight = 'bold')
```

## Male Population

### Row {height="50%"}

```{r}
#| label: male-projection-plot

# Get the appropriate forecast data for male population
male_data <- get_data_for_population("male")
forecast_data <- male_data$forecast_data
color_actual <- male_data$color
title <- "Actual (2010 Q1 - 2024 Q2) vs Forecasted Male Prison Population"

# Find the transition point between actual and forecast
actual_data <- forecast_data[!is.na(forecast_data$Actual),]
max_date <- max(actual_data$Quarter_Year)

# Get horizon dates for filtering
horizon_dates <- seq(max_date + 1, by = 1, length.out = projection_horizon)

# Create plot
p <- plot_ly()

# Add historical data
p <- p %>% add_lines(
  data = forecast_data[!is.na(forecast_data$Actual),],
  x = ~Quarter_Year,
  y = ~Actual,
  name = "Actual",
  line = list(color = color_actual)
)

# Add forecast data
forecast_part <- forecast_data[forecast_data$Quarter_Year %in% horizon_dates,]
p <- p %>% add_lines(
  data = forecast_part,
  x = ~Quarter_Year,
  y = ~Forecasted,
  name = "Forecasted",
  line = list(color = "#fdbf11", dash = "dash")
)

# Complete layout
p %>% layout(
  title = title,
  xaxis = list(
    title = "Quarter/Year",
    tickformat = "%Y Q%q",
    tickmode = "array",
    tickvals = c(
      yearquarter("2010 Q1"),
      yearquarter("2015 Q1"),
      yearquarter("2020 Q1"),
      yearquarter("2025 Q1"),
      yearquarter("2030 Q1")
    ),
    ticktext = c("2010 Q1", "2015 Q1", "2020 Q1", "2025 Q1", "2030 Q1")
  ),
  yaxis = list(title = "Prison Population", tickformat = ","),
  hovermode = "x unified",
  showlegend = TRUE,
  legend = list(title = list(text = "Series"))
)
```

### Row {height="30%"}

```{r}
#| label: male-projection-table

# Get the correct table data for male population
table_data <- male_data$table_data

# Limit rows to selected horizon
table_data <- head(table_data, projection_horizon)

# Format table
table_data <- table_data %>%
  mutate(
    Quarter_Year = format(Quarter_Year, "%Y Q%q"),
    Forecasted_hybrid = round(Forecasted_hybrid, 0)
  ) %>%
  select(
    Quarter_Year,
    Forecasted_hybrid,
    Percent_Change
  )

datatable(
  table_data,
  options = list(
    pageLength = 5,
    dom = 't',
    ordering = FALSE
  ),
  colnames = c('Quarter Year' = 'Quarter_Year', 
               'Forecasted' = 'Forecasted_hybrid', 
               '% Change' = 'Percent_Change'),
  rownames = FALSE
) %>%
  formatStyle(
    'Percent_Change',
    color = styleInterval(c(0), c('red', 'green')),
    fontWeight = 'bold'
  ) %>%
  formatRound('Forecasted_hybrid', 0) %>%
  formatRound('Percent_Change', 1) %>%
  formatStyle('Forecasted_hybrid', fontWeight = 'bold')
```

## Female Population

### Row {height="50%"}

```{r}
#| label: female-projection-plot

# Get the appropriate forecast data for female population
female_data <- get_data_for_population("female")
forecast_data <- female_data$forecast_data
color_actual <- female_data$color
title <- "Actual (2010 Q1 - 2024 Q2) vs Forecasted Female Prison Population"

# Find the transition point between actual and forecast
actual_data <- forecast_data[!is.na(forecast_data$Actual),]
max_date <- max(actual_data$Quarter_Year)

# Get horizon dates for filtering
horizon_dates <- seq(max_date + 1, by = 1, length.out = projection_horizon)

# Create plot
p <- plot_ly()

# Add historical data
p <- p %>% add_lines(
  data = forecast_data[!is.na(forecast_data$Actual),],
  x = ~Quarter_Year,
  y = ~Actual,
  name = "Actual",
  line = list(color = color_actual)
)

# Add forecast data
forecast_part <- forecast_data[forecast_data$Quarter_Year %in% horizon_dates,]
p <- p %>% add_lines(
  data = forecast_part,
  x = ~Quarter_Year,
  y = ~Forecasted,
  name = "Forecasted",
  line = list(color = "#fdbf11", dash = "dash")
)

# Complete layout
p %>% layout(
  title = title,
  xaxis = list(
    title = "Quarter/Year",
    tickformat = "%Y Q%q",
    tickmode = "array",
    tickvals = c(
      yearquarter("2010 Q1"),
      yearquarter("2015 Q1"),
      yearquarter("2020 Q1"),
      yearquarter("2025 Q1"),
      yearquarter("2030 Q1")
    ),
    ticktext = c("2010 Q1", "2015 Q1", "2020 Q1", "2025 Q1", "2030 Q1")
  ),
  yaxis = list(title = "Prison Population", tickformat = ","),
  hovermode = "x unified",
  showlegend = TRUE,
  legend = list(title = list(text = "Series"))
)
```

### Row {height="30%"}

```{r}
#| label: female-projection-table

# Get the correct table data for female population
table_data <- female_data$table_data

# Limit rows to selected horizon
table_data <- head(table_data, projection_horizon)

# Format table
table_data <- table_data %>%
  mutate(
    Quarter_Year = format(Quarter_Year, "%Y Q%q"),
    Forecasted_hybrid = round(Forecasted_hybrid, 0)
  ) %>%
  select(
    Quarter_Year,
    Forecasted_hybrid,
    Percent_Change
  )

datatable(
  table_data,
  options = list(
    pageLength = 5,
    dom = 't',
    ordering = FALSE
  ),
  colnames = c('Quarter Year' = 'Quarter_Year', 
               'Forecasted' = 'Forecasted_hybrid', 
               '% Change' = 'Percent_Change'),
  rownames = FALSE
) %>%
  formatStyle(
    'Percent_Change',
    color = styleInterval(c(0), c('red', 'green')),
    fontWeight = 'bold'
  ) %>%
  formatRound('Forecasted_hybrid', 0) %>%
  formatRound('Percent_Change', 1) %>%
  formatStyle('Forecasted_hybrid', fontWeight = 'bold')
```

### Row {height="20%"}

::: callout-note
For detailed data for these projections, see the [technical workbook](https://justice.utah.gov/research-analysis/).
:::


# ↕️️ Prison Drivers

## Gender & Offense Type

```{r}
#| label: prison-drivers
#| warning: false

# Use the pre-processed driver data that was loaded
p <- df_driver_1 |> 
  filter(!is.na(gender), !is.na(top_ofnse_typ_desc)) |>
  mutate(Gender = as.character(gender)) |>
  ggplot(aes(x = Quarter_Year, y = Count,
             group = Gender, colour=Gender)) +
  stat_summary(fun = sum, geom = "line") +
  labs(title = "Utah prison population (by offense type & gender)",
       y = "Number of prisoners") +
  facet_wrap(~ as.character(top_ofnse_typ_desc),
             nrow = 4, scales = "free_y") +
  theme(axis.text.x = element_text(angle = 90)) +
  labs(y = "Number of prisoners",
       x = "Year/Quarter",
       title = "Utah Prison Population (by gender and offense type): 2010:Q1-Present") + 
  scale_color_manual(values = c("#00A7C4", "#DA9100", "grey"), 
                     labels = c("Female", "Male", "Unknown")) + 
  guides(color = guide_legend(title = "Gender")) + 
  scale_x_yearquarter(breaks = seq(yearquarter("2010 Q1"), yearquarter("2025 Q4"), by = 8),
                     labels = ~format(.x, "%Y Q%q")) + 
  theme(axis.text.x = element_text(angle = 90, hjust = 1))
p <- p + geom_vline(xintercept = yearquarter("2020 Q1"), linetype = "dashed", color = "grey")  # Add vertical line

ggplotly(p)
```


# ⚠️ Capacity {.tabset}

## Total Population

```{r}
#| label: total-capacity-analysis

# Get the appropriate forecast data for total population
total_data <- get_data_for_population("total")
forecast_data <- total_data$forecast_data
color_actual <- total_data$color
title <- paste(total_data$title, "vs. Capacity")
capacity_limit <- total_data$capacity

# Find the transition point between actual and forecast
actual_data <- forecast_data[!is.na(forecast_data$Actual),]
max_date <- max(actual_data$Quarter_Year)

# Get horizon dates for filtering
horizon_dates <- seq(max_date + 1, by = 1, length.out = projection_horizon)

# Create capacity dataframe
capacity_df <- data.frame(
  Quarter_Year = unique(forecast_data$Quarter_Year),
  Capacity = capacity_limit
)

# Calculate when capacity is reached
capacity_reached <- NA
for (i in 1:length(horizon_dates)) {
  date <- horizon_dates[i]
  projected_value <- forecast_data$Forecasted[forecast_data$Quarter_Year == date]
  if (!is.na(projected_value) && projected_value > capacity_limit) {
    capacity_reached <- date
    break
  }
}

# Create plot
p <- plot_ly() %>%
  # Historical data
  add_lines(
    data = forecast_data[!is.na(forecast_data$Actual),],
    x = ~Quarter_Year,
    y = ~Actual,
    name = "Historical",
    line = list(color = color_actual)
  )

# Add forecasted data for the selected horizon
forecast_part <- forecast_data[forecast_data$Quarter_Year %in% horizon_dates,]
p <- p %>% add_lines(
  data = forecast_part,
  x = ~Quarter_Year,
  y = ~Forecasted,
  name = "Forecasted",
  line = list(color = "#fdbf11", dash = "dash")
)

# Add capacity line
p <- p %>% add_lines(
  data = capacity_df,
  x = ~Quarter_Year,
  y = ~Capacity,
  name = "Capacity Limit",
  line = list(color = "red", dash = "dot", width = 2)
)

# Add capacity reached annotation if applicable
if (!is.na(capacity_reached)) {
  projected_value <- forecast_data$Forecasted[forecast_data$Quarter_Year == capacity_reached]
  p <- p %>% add_annotations(
    x = capacity_reached,
    y = projected_value,
    text = paste("Capacity reached:", format(capacity_reached, "%Y Q%q")),
    showarrow = TRUE,
    arrowhead = 1,
    arrowsize = 1,
    arrowwidth = 2,
    arrowcolor = "red",
    ax = 0,
    ay = -40,
    font = list(color = "red")
  )
}

# Complete the plot
p %>% layout(
  title = title,
  xaxis = list(
    title = "Quarter/Year",
    tickformat = "%Y Q%q",
    tickmode = "array",
    tickvals = c(
      yearquarter("2010 Q1"),
      yearquarter("2015 Q1"),
      yearquarter("2020 Q1"),
      yearquarter("2025 Q1"),
      yearquarter("2030 Q1")
    ),
    ticktext = c("2010 Q1", "2015 Q1", "2020 Q1", "2025 Q1", "2030 Q1")
  ),
  yaxis = list(title = "Prison Population", tickformat = ","),
  hovermode = "x unified",
  showlegend = TRUE
)
```

## Male Population

```{r}
#| label: male-capacity-analysis

# Get the appropriate forecast data for male population
male_data <- get_data_for_population("male")
forecast_data <- male_data$forecast_data
color_actual <- male_data$color
title <- paste(male_data$title, "vs. Capacity")
capacity_limit <- male_data$capacity

# Find the transition point between actual and forecast
actual_data <- forecast_data[!is.na(forecast_data$Actual),]
max_date <- max(actual_data$Quarter_Year)

# Get horizon dates for filtering
horizon_dates <- seq(max_date + 1, by = 1, length.out = projection_horizon)

# Create capacity dataframe
capacity_df <- data.frame(
  Quarter_Year = unique(forecast_data$Quarter_Year),
  Capacity = capacity_limit
)

# Calculate when capacity is reached
capacity_reached <- NA
for (i in 1:length(horizon_dates)) {
  date <- horizon_dates[i]
  projected_value <- forecast_data$Forecasted[forecast_data$Quarter_Year == date]
  if (!is.na(projected_value) && projected_value > capacity_limit) {
    capacity_reached <- date
    break
  }
}

# Create plot
p <- plot_ly() %>%
  # Historical data
  add_lines(
    data = forecast_data[!is.na(forecast_data$Actual),],
    x = ~Quarter_Year,
    y = ~Actual,
    name = "Historical",
    line = list(color = color_actual)
  )

# Add forecasted data for the selected horizon
forecast_part <- forecast_data[forecast_data$Quarter_Year %in% horizon_dates,]
p <- p %>% add_lines(
  data = forecast_part,
  x = ~Quarter_Year,
  y = ~Forecasted,
  name = "Forecasted",
  line = list(color = "#fdbf11", dash = "dash")
)

# Add capacity line
p <- p %>% add_lines(
  data = capacity_df,
  x = ~Quarter_Year,
  y = ~Capacity,
  name = "Capacity Limit",
  line = list(color = "red", dash = "dot", width = 2)
)

# Add capacity reached annotation if applicable
if (!is.na(capacity_reached)) {
  projected_value <- forecast_data$Forecasted[forecast_data$Quarter_Year == capacity_reached]
  p <- p %>% add_annotations(
    x = capacity_reached,
    y = projected_value,
    text = paste("Capacity reached:", format(capacity_reached, "%Y Q%q")),
    showarrow = TRUE,
    arrowhead = 1,
    arrowsize = 1,
    arrowwidth = 2,
    arrowcolor = "red",
    ax = 0,
    ay = -40,
    font = list(color = "red")
  )
}

# Complete the plot
p %>% layout(
  title = title,
  xaxis = list(
    title = "Quarter/Year",
    tickformat = "%Y Q%q",
    tickmode = "array",
    tickvals = c(
      yearquarter("2010 Q1"),
      yearquarter("2015 Q1"),
      yearquarter("2020 Q1"),
      yearquarter("2025 Q1"),
      yearquarter("2030 Q1")
    ),
    ticktext = c("2010 Q1", "2015 Q1", "2020 Q1", "2025 Q1", "2030 Q1")
  ),
  yaxis = list(title = "Prison Population", tickformat = ","),
  hovermode = "x unified",
  showlegend = TRUE
)
```

## Female Population

```{r}
#| label: female-capacity-analysis

# Get the appropriate forecast data for female population
female_data <- get_data_for_population("female")
forecast_data <- female_data$forecast_data
color_actual <- female_data$color
title <- paste(female_data$title, "vs. Capacity")
capacity_limit <- female_data$capacity

# Find the transition point between actual and forecast
actual_data <- forecast_data[!is.na(forecast_data$Actual),]
max_date <- max(actual_data$Quarter_Year)

# Get horizon dates for filtering
horizon_dates <- seq(max_date + 1, by = 1, length.out = projection_horizon)

# Create capacity dataframe
capacity_df <- data.frame(
  Quarter_Year = unique(forecast_data$Quarter_Year),
  Capacity = capacity_limit
)

# Calculate when capacity is reached
capacity_reached <- NA
for (i in 1:length(horizon_dates)) {
  date <- horizon_dates[i]
  projected_value <- forecast_data$Forecasted[forecast_data$Quarter_Year == date]
  if (!is.na(projected_value) && projected_value > capacity_limit) {
    capacity_reached <- date
    break
  }
}

# Create plot
p <- plot_ly() %>%
  # Historical data
  add_lines(
    data = forecast_data[!is.na(forecast_data$Actual),],
    x = ~Quarter_Year,
    y = ~Actual,
    name = "Historical",
    line = list(color = color_actual)
  )

# Add forecasted data for the selected horizon
forecast_part <- forecast_data[forecast_data$Quarter_Year %in% horizon_dates,]
p <- p %>% add_lines(
  data = forecast_part,
  x = ~Quarter_Year,
  y = ~Forecasted,
  name = "Forecasted",
  line = list(color = "#fdbf11", dash = "dash")
)

# Add capacity line
p <- p %>% add_lines(
  data = capacity_df,
  x = ~Quarter_Year,
  y = ~Capacity,
  name = "Capacity Limit",
  line = list(color = "red", dash = "dot", width = 2)
)

# Add capacity reached annotation if applicable
if (!is.na(capacity_reached)) {
  projected_value <- forecast_data$Forecasted[forecast_data$Quarter_Year == capacity_reached]
  p <- p %>% add_annotations(
    x = capacity_reached,
    y = projected_value,
    text = paste("Capacity reached:", format(capacity_reached, "%Y Q%q")),
    showarrow = TRUE,
    arrowhead = 1,
    arrowsize = 1,
    arrowwidth = 2,
    arrowcolor = "red",
    ax = 0,
    ay = -40,
    font = list(color = "red")
  )
}

# Complete the plot
p %>% layout(
  title = title,
  xaxis = list(
    title = "Quarter/Year",
    tickformat = "%Y Q%q",
    tickmode = "array",
    tickvals = c(
      yearquarter("2010 Q1"),
      yearquarter("2015 Q1"),
      yearquarter("2020 Q1"),
      yearquarter("2025 Q1"),
      yearquarter("2030 Q1")
    ),
    ticktext = c("2010 Q1", "2015 Q1", "2020 Q1", "2025 Q1", "2030 Q1")
  ),
  yaxis = list(title = "Prison Population", tickformat = ","),
  hovermode = "x unified",
  showlegend = TRUE
)

```


# 📝 Methodology

## Row

```{r}
HTML('
<div class="methodology-card">
  <h3>Prison Population Projection Methodology</h3>
  
  <h4>Hybrid Forecasting Approach</h4>
  <p>The projections in this dashboard use a hybrid modeling approach that combines:</p>
  
  <ol>
    <li><strong>Linear Regression with Structural Breaks:</strong> 
      The base model captures long-term trends and policy impacts, including:
      <ul>
        <li>Time trend component</li>
        <li>JRI (Justice Reinvestment Initiative) policy impact (2015)</li>
        <li>COVID-19 impact (2020)</li>
      </ul>
    </li>
    <li><strong>Exponential Smoothing (ETS):</strong> 
      Applied to the residuals of the linear model to capture short-term patterns, with alpha parameter set to 0.8 to emphasize recent patterns
    </li>
  </ol>
  
  <h4>Data Sources</h4>
  <p>The projections are based on historical prison population data by demographic groups, analyzed at the quarterly level to balance detail with forecasting stability.</p>
</div>
')
```

# Additional Information {.sidebar-toggle}

## Limitations

```{r}
HTML('
<div class="limitations-card">
  <h4>Projection Limitations</h4>
  <p>These projections assume continuation of current trends and policies. Unexpected policy changes or events could significantly alter outcomes.</p>
  
  <h4>Model Considerations</h4>
  <ul>
    <li>The model accounts for known structural breaks (JRI and COVID) but cannot anticipate future policy shifts</li>
    <li>Quarterly data balances detail with forecasting stability</li>
    <li>Models are regularly evaluated and recalibrated as new data becomes available</li>
  </ul>
</div>
')
```

## Data Sources

```{r}
HTML('
<div class="data-sources-card">
  <h4>Primary Data Sources</h4>
  <ul>
    <li>Utah Department of Corrections (UDC) - Offender Management System</li>
    <li>Commission on Criminal and Juvenile Justice (CCJJ) - Annual Reports</li>
    <li>Bureau of Justice Statistics (BJS) - Annual Prison Population Survey</li>
  </ul>
  
  <h4>Methodology References</h4>
  <p>For detailed information on the forecasting methodology, please see the <a href="https://justice.utah.gov/research-analysis/">technical workbook</a>.</p>
</div>
')
```

## Custom CSS

```{css}
.methodology-card, .limitations-card, .data-sources-card {
  background-color: #f8f9fa;
  border-radius: 5px;
  padding: 15px;
  margin-bottom: 20px;
  border-left: 4px solid #00A7C4;
}

.limitations-card {
  border-left-color: #fdbf11;
}

.data-sources-card {
  border-left-color: #0a4c6a;
}

.methodology-card h3, .limitations-card h4, .data-sources-card h4 {
  color: #333;
  margin-top: 0;
}

.methodology-card ul, .limitations-card ul, .data-sources-card ul {
  padding-left: 20px;
}
```


  